---
title: "FAST: dorsolateral_prefrontal_cortex"
author: "Wei Liu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAST: dorsolateral_prefrontal_cortex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


This Rmd file introduce how to apply FAST to obtain the embeddings for the dorsolateral prefrontal cortex Visium data. This is corresponding the `dorsolateral_prefrontal_cortex.R` file.


## Source the required functions
```{r  eval=TRUE}
source("util_funcs.R")
```


## Start anslysis
Load the required packages.
```{r  eval=TRUE}
library(ProFAST)
library(SingleCellExperiment)
library(Seurat)
library(Matrix)
library(dplyr)
```


## Load data and preprocessing ---------------------------------------------

```{r  eval=TRUE}
name_ID12 <- as.character(c(151507, 151508, 151509, 151510, 151669, 151670,
                            151671, 151672, 151673, 151674, 151675, 151676))


n_sample <- length(name_ID12)
url_brainA <- "https://github.com/feiyoung/DR-SC.Analysis/raw/main/data/DLPFC_data/"; url_brainB <- ".rds"
HVGList <- list()
timeVec <- rep(NA, n_sample)
seuList_filter <- list()
for(r in 1:n_sample){
  # r <- 1
  
  cat('input brain data', r, '\n')
  # load and read data
  dlpfc <- readRDS(url(paste0(url_brainA, name_ID12[r],url_brainB) ))
  
  sp_count <- dlpfc@assays@data$counts
  meta.data <- as.data.frame(colData(dlpfc))
  ## Filter counts:
  min_spots <- 20
  gene_flag <- rowSums(sp_count>0)>min_spots
  sp_count_filter <-  sp_count[gene_flag, ]
  min_genes <- 20
  spot_flag <- colSums(sp_count>0) > min_genes
  sp_count_filter <- sp_count_filter[, spot_flag]
  
  tic <- proc.time()
  seu <- CreateSeuratObject(counts=sp_count_filter, meta.data = meta.data[spot_flag,])
  seu <-  NormalizeData(seu)
  seu <- FindVariableFeatures(seu)
  toc <- proc.time()
  var.features <- seu@assays$RNA@var.features
  seuList_filter[[r]] <- seu
  HVGList[[r]] <- var.features
  timeVec[r] <- toc[3] - tic[3]
}

gene_intersect <- Reduce(intersect, lapply(seuList_filter, row.names))
## transfer to symbols
gene_symbols <- transferGeneNames(gene_intersect, species="Human")
gene_symbols <- toupper(gene_symbols)
names(gene_symbols) <- gene_intersect

```


Select the common variable genes
```{r  eval=TRUE}
## select the integrated HVGs
gene_hvg_2000 <- selectIntFeatures(seuList_filter,HVGList) # 

```

## Get Log-nomalized data list

```{r  eval=TRUE}

seulist_hvg <- lapply(seuList_filter, function(x) x[gene_hvg_2000, ])
yList <- lapply(seulist_hvg, function(x) x$layer_guess_reordered)
for(r in 1:length(yList)){
  y_true <- as.character(yList[[r]])
  y_true[is.na(y_true)] <- "NA"
  yList[[r]] <- y_true 
}
posList <- list()
AdjList <- list()
## To differetiate different sections by adding a large value for the coordinates
## since SpatialPCA and DR-SC, used the spatial coordinates, are designed to handle one section.
for(m in 1: length(seulist_hvg)){
  message("m = ", m)
  pos <- cbind(seulist_hvg[[m]]$row, seulist_hvg[[m]]$col)
  AdjList[[m]] <- PRECAST::getAdj_reg(pos, platform = "Visium")
  posList[[m]] <- pos + (m-1)* 1e6 
  
}

seulist_hvg <- lapply(seulist_hvg, NormalizeData)
XList1 <- lapply(seulist_hvg, function(x) Matrix::t(x[["RNA"]]@data))
sum(sapply(XList1, nrow))

```

## Fit FAST-P
First, we fit the Poisson version of FAST.
```{r  eval=TRUE}
hq <- 15

XList_count <- lapply(seulist_hvg, function(x) Matrix::t(x[["RNA"]]@counts))
tic <- proc.time()
reslist_profastP <- ProFAST_run(XList_count, AdjList = AdjList, fit.model = "poisson", q=hq)
toc <- proc.time()
time_profastP <- toc[3] - tic[3]
## save(reslist_profastP,time_profastP, file="reslist_hvg_profastP_DLPFC12.rds")

## calculate the  Macffaden's R2 for each section
(R2_profastP <- evaluate_DR_PF2(reslist_profastP$hV, yList))

```

## Fit FAST-G

Next, we fit FAST-G.
```{r  eval=TRUE}

### Gaussian version 
tic <- proc.time()
reslist_profastG <- ProFAST_run(XList1, AdjList = AdjList, q=hq)
toc <- proc.time()
time_profastG <- toc[3] - tic[3]

## calculate the  Macffaden's R2 for each section
(R2_profastG <- evaluate_DR_PF2(reslist_profastG$hV, yList))
```




## Performance
Output the Macffaden's R2 for each section.

```{r  eval=TRUE}
macR2 <- cbind(FAST_P=R2_profastP, FAST_G=R2_profastG)
print(macR2)
```


<details>
<summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>